---
title: "DEA w DESeq2 on the counts from recount2"
output:
  html_notebook:
    code_folding: show
editor_options:
  chunk_output_type: console
---
# {.tabset .tabset-fade .tabset-pills}

## Setup
### Load libraries
```{r}
library(here)
library(tidyverse)
library(tidylog)
library(conflicted)
conflict_prefer("select", "tidylog")
conflict_prefer("left_join", "tidylog")
conflict_prefer("filter_at", "tidylog")
conflict_prefer("semi_join", "tidylog")
conflict_prefer("distinct", "tidylog")
conflict_prefer("mutate", "tidylog")
conflict_prefer("filter_all", "tidylog")
conflict_prefer("filter", "tidylog")
```

### Download and read in the gene counts from recount2
```{r}
recount2_url <- "http://duffel.rail.bio/recount/v2/SRP019968/counts_gene.tsv.gz"
destfile <- here("data", "SRP019968_recount2_gene_counts.tsv.gz")

if (!file.exists(destfile)) {
  download.file(url = recount2_url, destfile = destfile)
}

recount2_counts <- read_tsv(here("data", "SRP019968_recount2_gene_counts.tsv.gz"),
                            col_names = T)
```

## Data prep
### Keep only protein-coding genes using biomaRt
```{r}
## Extract ensembl ids from recount2 data
recount2_counts <- recount2_counts %>% 
  ## remove duplicate genes that can be found either on chromosome X or Y
  filter_at(vars(gene_id), ~str_detect(., "PAR_Y", negate = T)) %>% 
  ## remove the version part of the Ensembl ids
  mutate(gene_id = str_remove(gene_id, "\\.\\w*$"))

recount2_ensg_ids <- recount2_counts %>% select(gene_id)

## Query the Ensembl Biomart for protein coding genes
hsapiens_gene_ensembl = biomaRt::useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")

protein_coding <- biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
                                 filters = c("ensembl_gene_id","biotype"),
                                 values=list(purrr::as_vector(recount2_ensg_ids),
                                             "protein_coding"),
                                 mart = hsapiens_gene_ensembl)


## Keep only the counts for protein coding genes
counts_pcoding <- recount2_counts %>% 
  semi_join(distinct(protein_coding, ensembl_gene_id), 
            by = c("gene_id" = "ensembl_gene_id"))
```

### How many protein-coding genes were detected?
```{r}
counts_pcoding %>% select(SRR837858, SRR837859) %>% filter_all(all_vars(. > 0)) %>% nrow() # 14941; 12493 in text
counts_pcoding %>% select(SRR837860, SRR837861) %>% filter_all(all_vars(. > 0)) %>% nrow() # 14540; 12173 in text
counts_pcoding %>% select(SRR837862, SRR837863) %>% filter_all(all_vars(. > 0)) %>% nrow() # 14548; 12246 in text
counts_pcoding %>% select(SRR837864, SRR837865) %>% filter_all(all_vars(. > 0)) %>% nrow() # 14543; 12074 in text
```

## DEA
### Prepare a count matrix for DESeq2
```{r}
count_matrix <- counts_pcoding %>% column_to_rownames("gene_id")
```

### Prepare column metadata for DESeq2
```{r}
col_data <- data.frame(row.names = colnames(count_matrix),
                            "lane" = rep(c("a", "b")),
                            "stage" = factor(rep(c("primary", "hTERT", "SV40", "RAS"),
                                                 each = 2)))

## Verify that col_data and count_matrix match
all(rownames(col_data) == colnames(count_matrix))
```

### Create DESeq object
```{r}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = count_matrix,
                                      colData = col_data,
                                      design = ~ lane + stage)

dds$stage <- factor(dds$stage, levels = c("primary", "hTERT", "SV40", "RAS"))
```

### Perform the DEA
```{r}
dds <- DESeq2::DESeq(dds)
```

### Extract the DEGs for metastasizing (RAS) vs immortalized (hTERT) fibroblasts contrast
```{r}
res4vs2 <- DESeq2::results(dds, contrast = c("stage", "RAS", "hTERT"), lfcThreshold = 1, alpha = 0.01)
DESeq2::summary(res4vs2)
     # 639 up-regulated, 1329 down-regulated, 1968 total

degs_4vs2 <- data.frame(res4vs2) %>% rownames_to_column("gene_id") %>% filter(padj < 0.01)
```

## Clean up
```{r}
degs_4vs2 %>% write_tsv(here("output", "degs_4vs2"))
varhandle::rm.all.but(keep = c("degs_4vs2", "dds"))
varhandle::save.var(path = here("analysis"), varlist = c("degs_4vs2", "dds"))
```


