---
title: "Network enrichment with NEArender"
output:
  html_notebook:
    code_folding: show
editor_options:
  chunk_output_type: console
---
# {.tabset .tabset-fade .tabset-pills}

## Setup
### Load libraries and resolve conflicts
```{r}
library(here)
library(tidyverse)
library(tidylog)
library(org.Hs.eg.db)
library(conflicted)
conflict_prefer("filter", "tidylog")
conflict_prefer("select", "tidylog")
conflict_prefer("drop_na", "tidylog")
conflict_prefer("filter_all", "tidylog")
conflict_prefer("full_join", "tidylog")
conflict_prefer("mutate", "tidylog")
conflict_prefer("distinct", "tidylog")
conflict_prefer("filter_at", "tidylog")
conflict_prefer("group_by", "tidylog")
conflict_prefer("rename_at", "tidylog")
```

## Prepare AGS with Entrez ids
```{r}
load(here("analysis", "20200120-205728" ,"all_degs_HPA.RData"))

## retrieve mapping
ags_ensg2entrez <- 
  all_degs_HPA$gene %>%
  AnnotationDbi::mapIds(keytype = "ENSEMBL",
                        column = "ENTREZID",
                        x = org.Hs.eg.db,
                        multiVals = "first") %>%
  enframe("ensg", "entrez_id") %>%
  filter(!duplicated(.$ensg))

n_distinct(all_degs_HPA$gene) # 730

## map the ids and construct the AGS object
ags_4vs2_entrez <- all_degs_HPA %>%
  full_join(ags_ensg2entrez, by = c("gene" = "ensg")) %>%
  as.data.frame() %>%
  NEArender::import.gs(gs.type = "a", col.gene = 5, col.set = 1, Lowercase = F)
```

## Prepare FGS with Entrez ids
### Reactome 
```{r}
## download and clean Reactome gene:pathway file with Entrez ids
url = "https://reactome.org/download/current/NCBI2Reactome_All_Levels.txt"
file = here("data", "NCBI2Reactome_All_Levels.txt")

if(!file.exists(file)){
  download.file(url = url, destfile = file)
}

reactome71_entrez <- 
  read_tsv(file = file,
           col_names = c("entrezId", "reactomePathwayId", "url", "eventName", "evidenceCode", "species"),
           col_types = cols("entrezId" = col_character())) %>% 
  filter(species == "Homo sapiens") %>%
  unite("pathway", reactomePathwayId, eventName, sep = "_") %>%
  select(entrezId, pathway)

n_distinct(reactome71_entrez$entrezId) # 10881
n_distinct(reactome71_entrez$pathway) # 2309

## create an FGS object
fgs_react71_entrez <- reactome71_entrez %>% 
  as.data.frame() %>%
  NEArender::import.gs(gs.type = "f", col.gene = 1, col.set = 2, Lowercase = F)

## Create a gmt file to use with Sumer's Affinity Propagation
react_gmt_3col <- reactome71_entrez %>%
  group_by(pathway) %>% 
  mutate(genes = str_c(entrezId, collapse = ",")) %>% 
  distinct(pathway, genes) %>% 
  separate(pathway, into = c("pathway_id", "pathway_name"), sep = "_", extra = "merge")
  
## Save the gmt file
react_gmt_3col %>%
  write_tsv(
    here("analysis", "full_gmts_entrez", "react_entrez_full.gmt"),
    na = "",
    col_names = F
)
```

### KEGG
```{r}
## Read and clean gmt file prepared by prepare_database function from geneSCF
## TODO: include the modified update_KEGG.sh script or rewrite it here 
kegg_gmt_3col <- read_tsv(here("data", "kegg_genescf_2020-01-08.gmt"),
                 col_names = c("pathway", "genes"),
                 col_types = cols(.default = "c")) %>% 
  separate(pathway, into = c("pathway_id", "pathway_name"), sep = "~") %>% 
  mutate(pathway_name = str_remove(pathway_name, fixed("_-_Homo_sapiens_(human)"))) %>% 
  mutate(pathway_name = str_replace_all(pathway_name, "_", " ")) 

nrow(kegg_gmt_3col) # 336 pathways

## Merge pathway ids with pathway names and cast the dataframe into a long format
kegg_long <- kegg_gmt_3col%>% 
  unite("pathway", pathway_id, pathway_name, sep = "_") %>% 
  splitstackshape::cSplit("genes", sep = ",", direction = "long", type.convert = F)

## Create an FGS object
fgs_kegg_entrez <- kegg_long %>%
  NEArender::import.gs(gs.type = "f", col.gene = 2, col.set = 1, Lowercase = F)

## Save the gmt file
kegg_gmt_3col %>% write_tsv(
  here("analysis", "full_gmts_entrez", "kegg_entrez_full.gmt"),
  na = "",
  col_names = F
)
```

### WikiPathways
```{r}
## download current gmt file from WP
file <- here("data", "wikipathways-20200110-gmt-Homo_sapiens.gmt")

if(!file.exists(file)){
  rWikiPathways::downloadPathwayArchive(organism = "Homo sapiens", format = "gmt", destpath = here("data"), date = "current") 
}

## clean the gmt file to contain 3 columns
wp_gmt_3col <- read_table(file, col_names = F) %>% 
  separate(X1, into = c("X1", "url", "genes"), sep = "\t", extra = "merge") %>% 
  separate(X1, into = c("pathway_name", "db_version", "pathway_id", "species"), sep = "%") %>% 
  select(pathway_name, pathway_id, genes) %>% 
  mutate(genes = str_replace_all(genes, pattern = "\t", replacement = ","))

n_distinct(wp_gmt_3col$pathway_id) # 565

## Merge pathway ids with pathway names and cast the dataframe into a long format
wp_long <- wp_gmt_3col %>% 
  unite("pathway", pathway_id, pathway_name, sep = "_") %>% 
  splitstackshape::cSplit("genes", sep = ",", direction = "long", type.convert = F)

## Create an FGS object
fgs_wp_entrez <- wp_long %>%
  NEArender::import.gs(gs.type = "f", col.set = 1, col.gene = 2, Lowercase = F)

## Create gmt file to use with Sumer
wp_gmt <- wp_gmt_3col %>% 
  splitstackshape::cSplit("genes", sep = ",", direction = "wide", type.convert = F)

## Save the gmt file
wp_gmt %>% write_tsv(
  here("analysis", "full_gmts_entrez", "wp_entrez_full.gmt"),
  na = "",
  col_names = F
)
```

## Prepare network with Entrez ids from PathwayCommons
```{r}
## Download PathwayCommons v12 SIF file
url <- "https://www.pathwaycommons.org/archives/PC2/v12/PathwayCommons12.All.hgnc.sif.gz"
file <- here("data", "PathwayCommons12.All.hgnc.sif.gz")

if(!file.exists(file)){
  download.file(url = url, destfile = file)
}

## Read and clean the SIF file
pc12 <-
  read_tsv(file = file,
           col_names = c("geneA", "interaction_type", "geneB")) %>%
  select(-interaction_type) %>% # get rid of the interaction type column
  filter_all(all_vars(str_starts(., pattern = "CHEBI:", negate = T))) # get rid of CHEBI symbols

nrow(pc12) # 1200159

## Check if the gene symbols are approved HGNC symbols, if not, retrieve the approved ones
approvedA <- pc12$geneA %>% 
  HGNChelper::checkGeneSymbols() %>% 
  splitstackshape::cSplit("Suggested.Symbol", sep = "///") %>% 
  select("Suggested.Symbol_1")

approvedB <- pc12$geneB %>% 
  HGNChelper::checkGeneSymbols() %>% 
  splitstackshape::cSplit("Suggested.Symbol", sep = "///") %>% 
  select("Suggested.Symbol_1")

pc12_approved_symbols <- bind_cols(approvedA, approvedB) %>% 
  setNames(c("geneA", "geneB")) %>% 
  drop_na() # CRIPAK and SPHAR are deprecated gene symbols; remove all of their interactions n=13
    
## Extract all uniqe gene symbols to be mapped to Entrez ids
pc12_symbols <- list("a" = as.character(pc12_approved_symbols$geneA),
                     "b" = as.character(pc12_approved_symbols$geneB)) %>% 
  stack() %>%
  distinct(values) # There are 19073 unique approved gene symbols in the WikiPathways SIF file

## Map gene symbols using org.Hs.eg.db
library("org.Hs.eg.db")
mapping_orgHs <- pc12_symbols %>%
  purrr::as_vector() %>% 
  AnnotationDbi::mapIds(keytype = "SYMBOL",
                        column = "ENTREZID",
                        x = org.Hs.eg.db,
                        multiVals = "first") %>% 
  enframe("gene_symbol", "entrez_id")

## Extract the approved gene symbols that did not map using org.Hs.eg.db and map them using biomaRt
not_mapped <- mapping_orgHs %>% 
  filter(is.na(entrez_id)) %>% # 28 gene symbols were not mapped by org.Hs.eg.db
  select(gene_symbol) %>%
  as_vector() %>%
  as.character()

ensembl_hs <- biomaRt::useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
additional_mapping <- not_mapped %>% 
  biomaRt::getBM(
    attributes = c("hgnc_symbol", "entrezgene_id"),
    filters = "hgnc_symbol",
    mart = ensembl_hs)

## One gene did not map even though it can be found in NCBI. Substitute it manually.
additional_mapping[additional_mapping$hgnc_symbol == "MT-RNR2", 2] <- "4550"

## Combine all of the mapping data into one dataframe
colnames(additional_mapping) <- colnames(mapping_orgHs)

all_mapping <- bind_rows(mapping_orgHs, additional_mapping) %>% 
  drop_na() # removed 28 rows (<1%), 19,059 rows remaining

## BONUS: Extract gene symbols that were not mapped by neither method
pc12_symbols %>% filter(!values %in% all_mapping$gene_symbol) # 14 gene symbols were not mapped

## Substitute the gene symbols in the WikiPathways SIF dataframe to Entrez ids
substituteA <- match(pc12_approved_symbols$geneA, all_mapping$gene_symbol) # match vector for geneA column
substituteB <- match(pc12_approved_symbols$geneB, all_mapping$gene_symbol) # match vector for geneB column

pc12A_entrez <- all_mapping$entrez_id[substituteA] %>% tibble::enframe() # map gene A
pc12B_entrez <- all_mapping$entrez_id[substituteB] %>% tibble::enframe() # map gene B

## Bind columns and filter out genes that were not mapped
pc12_entrez <- bind_cols(pc12A_entrez, pc12B_entrez) %>%
  select(value, value1) %>% 
  filter_all(all_vars(!is.na(.)))
    # removed 711 rows (<1%), 1,199,435 rows remaining (out of 1200159 original interactions in the SIF file)

## construct the NET object
net_pc12_entrez <- pc12_entrez %>% 
  as.data.frame() %>%
  NEArender::import.net(Lowercase = F)
    # "Network of 1038803 edges between 19059 nodes..."
```

## Save the AGS, NET, and 3 FGS objects
```{r}
variables_to_save <- c("ags_4vs2_entrez",
                       "net_pc12_entrez",
                       "fgs_kegg_entrez",
                       "fgs_react71_entrez",
                       "fgs_wp_entrez")

varhandle::save.var(path = here("analysis"), varlist = variables_to_save, newdirtag = "NEArender_input")
varhandle::rm.all.but(keep = variables_to_save)
```

## Perfom NEA
```{r}
nea_react <- NEArender::nea.render(AGS = ags_4vs2_entrez,
                                   FGS = fgs_react71_entrez,
                                   NET = net_pc12_entrez,
                                   Parallelize = 8,
                                   members = T,
                                   graph = T)

nea_kegg <- NEArender::nea.render(AGS = ags_4vs2_entrez,
                                  FGS = fgs_kegg_entrez,
                                  NET = net_pc12_entrez,
                                  Parallelize = 8,
                                  members = T,
                                  graph = T)

nea_wp <- NEArender::nea.render(AGS = ags_4vs2_entrez,
                                FGS = fgs_wp_entrez,
                                NET = net_pc12_entrez,
                                Parallelize = 8,
                                members = T,
                                graph = T)

varhandle::save.var(newdirtag = "NEArender_raw_results",
                    path = here("analysis"),
                    varlist = c("nea_kegg", "nea_react", "nea_wp"))
```

## Extract results
### Define functions
#### extractNeaResult()
```{r}
extractNeaResults <- function(nea_results) {
  members_ags <- nea_results$members.ags %>%
    t.default() %>%
    as.data.frame.matrix() %>%
    rownames_to_column("pathway") %>%
    rename_at(.vars = vars(-pathway), .funs = ~paste0("ags_", .))
  
  members_fgs <- nea_results$members.fgs %>%
    as.data.frame.matrix() %>%
    rownames_to_column("pathway") %>%
    rename_at(vars(-pathway), ~paste0("fgs_", .))

  q_values <- nea_results$q %>%
    as.data.frame.matrix() %>%
    rownames_to_column("pathway") %>%
    rename_at(vars(-pathway), ~paste0("qval_", .))
  
  n_missing_qvals <- is.na(q_values) %>% sum()
  print(paste("Number of missing q values: ", n_missing_qvals))
  
  nea_results_clean <- q_values %>%
    full_join(members_ags, by = "pathway") %>%
    full_join(members_fgs, by = "pathway") %>%
    separate(pathway, into = c("pathway_id", "pathway_name"), sep = "_", extra = "merge")

  nea_results_clean
}
```

#### significantPerCancer()
```{r}
significantPerCancer <- function(nea_results ,cancer, threshold) {
  nea_results %>%
    select(starts_with("pathway"), contains(cancer)) %>%
    filter_at(vars(contains("qval")), any_vars(. < threshold))
}
```

### Clean the results and extract significant pathways
#### WikiPathways
```{r}
res_nea_wp <- nea_wp %>% extractNeaResults()
brca_wp <- res_nea_wp %>% significantPerCancer("brca", 0.01) # 130; 0
luad_wp <- res_nea_wp %>% significantPerCancer("luad", 0.01) # 218; 1
prad_wp <- res_nea_wp %>% significantPerCancer("prad", 0.01) # 243; 1
coadread_wp <- res_nea_wp %>% significantPerCancer("coadread", 0.01) # 110; 1
```

#### KEGG
```{r}
res_nea_kegg <- nea_kegg %>% extractNeaResults()
brca_kegg <- res_nea_kegg %>% significantPerCancer("brca", 0.01) # 133; 193
luad_kegg <- res_nea_kegg %>% significantPerCancer("luad", 0.01) # 134; 132
prad_kegg <- res_nea_kegg %>% significantPerCancer("prad", 0.01) # 182; 184
coadread_kegg <- res_nea_kegg %>% significantPerCancer("coadread", 0.01) # 100; 187
```

#### Reactome
```{r}
res_nea_react <- nea_react %>% extractNeaResults()
brca_react <- res_nea_react %>% significantPerCancer("brca", 0.01) # 281; 486
luad_react <- res_nea_react %>% significantPerCancer("luad", 0.01) # 516; 253
prad_react <- res_nea_react %>% significantPerCancer("prad", 0.01) # 395; 364
coadread_react <- res_nea_react %>% significantPerCancer("coadread", 0.01) # 242; 322
```

## Explore the results
### KEGG
```{r}
venn_kegg <- gplots::venn(
  list(
    "BRCA" = brca_kegg$pathway_name,
    "COADREAD" = coadread_kegg$pathway_name,
    "LUAD" = luad_kegg$pathway_name,
    "PRAD" = prad_kegg$pathway_name
  )
)
attr(venn_kegg, "intersections")

bind_rows(brca_kegg, coadread_kegg, prad_kegg, luad_kegg) %>% distinct(pathway_name) %>% count() # 220
```

### WikiPathways
```{r}
venn_wp <- gplots::venn(
  list(
    "BRCA" = brca_wp$pathway_name,
    "COADREAD" = coadread_wp$pathway_name,
    "LUAD" = luad_wp$pathway_name,
    "PRAD" = prad_wp$pathway_name
  )
)
attr(venn_wp, "intersections")

bind_rows(brca_wp, coadread_wp, prad_wp, luad_wp) %>% distinct(pathway_name) %>% count() # 294
```

### Reactome
```{r}
venn_react <- gplots::venn(
  list(
    "BRCA" = brca_react$pathway_name,
    "COADREAD" = coadread_react$pathway_name,
    "LUAD" = luad_react$pathway_name,
    "PRAD" = prad_react$pathway_name
  )
)
attr(venn_react, "intersections")

bind_rows(brca_react, coadread_react, prad_react, luad_react) %>% distinct(pathway_name) %>% count()
```

## Export results
### remove unnecessary variables and save the rest
```{r}
varhandle::rm.all.but(keep = c("brca*", "prad*", "luad*", "coadread*", "*gmt_3col"), keep_functions = F)
varhandle::save.var(path = here("output_data"), newdirtag = "from_NEArender")
```
